Four decades of water quality change in the upper San Francisco Estuary

### 2017-05-02 sftrends_manu.tex
slight shortening

Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the northern \ac{sfe} region.  The \ac{iep} is a consortium of state and federal agencies that have maintained the \ac{emp} in the Delta region since 1975 \cite{IEP13}.  The \ac{emp} collects monthly water quality samples at 19 stations in the Delta, Suisun Bay, and northeastern San Pablo Bay.  Water samples were collected using a Van Dorn sample, a submersible pump, or a flow through system depending on site.  All samples were processed with standard QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \cite{IEP13}.  Nutrient time series from 1976 to 2013 were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) for ten discrete sampling stations (\cref{fig:delt_map}).  Stations were grouped by location in the study area for comparison: peripheral Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; interior Delta stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations were chosen based on continuity of the water quality time series and significance of their geographic location for understanding regional trends.  Time series were complete for all stations except for an approximate ten year gap from 1996-2014 for D19.  Data were minimally processed, with the exception of averaging replicates that occurred on the same day.  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two). Less than 3\% of all observations were left-censored, although variation was observed between analytes and location.  The ammonium time series had the most censored observations at sites C10 (25.4\% of all observations), MD10 (18.1\%), D28 (17.8\%), D19 (12\%), and D7 (7.9\%).

Daily flow estimates for the Delta region were obtained from the Dayflow software program that provides estimates of average Delta outflow \cite{IEP16}.  Because of the complexity of water inflow, exports, and outflows from the Delta, the Dayflow program combines observations with estimates based on mass balance to reconstruct historical and daily flow estimates.  The \ac{wrtds} models described below require a matched flow record with the appropriate station to evaluate nutrient trends. Given the complexity of inflows and connectivity of the system, only the inflow estimates from the Sacramento and San Joaquin rivers were used as measures of freshwater influence at each station.  Initial analyses indicated that model fit was not significantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series.  As such, the Sacramento daily flow time series was used to account for flow effects at C3, D19, D26, D28, and MD10, and the San Joaquin time series was used for C10 and P8 based on station proximity to each inflow.  Salinity observations at D4, D6, and D7 in Suisun Bay were used as more appropriate measures of freshwater variation, given the stronger tidal influence at these stations.  Salinity has been used as a tracer of freshwater influence for the application of \ac{wrtds} models in tidal waters \cite{Beck15}.

@@

The observed time series for the ten Delta - Suisun Bay stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).  In general, long-term (inter-annual) trends were not easily observed from the raw data. \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.  However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick15}.  By location, observed concentrations of \ac{din} for the entire time series were higher on average for the peripheral stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: $1.04$$\pm$$0.03$ \mgl) and similar for the interior (D19, D26, D28, $0.43$$\pm$$0.01$) and Suisun Bay stations (D4, D6, D7, $0.44$$\pm$$0.01$).  Average concentrations were highest at P8 ($1.63$$\pm$$0.05$ \mgl) and lowest at C3 ($0.4$$\pm$$0.01$) for \ac{din}, highest at P8 ($0.28$$\pm$$0.02$) and lowest at D28 ($0.05$$\pm$$0.003$) for ammonium, and highest at C10 ($1.4$$\pm$$0.04$) and lowest at C3 ($0.15$$\pm$$0.004$) for nitrite/nitrate. Mean observed concentrations were also higher later in the time series for all analytes.  For example, average \ac{din} across all stations was $0.61$$\pm$$0.01$ \mgl for 1976-1995, compared to $0.7$$\pm$$0.01$ for 1996-2013. Seasonal changes across all years showed that nitrogen concentrations were generally lower in the summer and higher in the winter, although observed patterns were inconsistent between sites.  For example, site MD10 had distinct seasonal spikes for elevated \ac{din} in the winter, whereas other stations had less prominent seasonal maxima (e.g., C3, D7, \cref{fig:obsdat}).

@@

Application of seasonal Kendall tests to evaluate trends in observed data provided explicit information on the direction, magnitude, and statistical significance of changes between years. Trends estimated from the observed data for 1976-1995 and 1996-2013 varied considerably between sites and analytes (\cref{fig:trndcomp1}). Significant trends were observed from 1976-1995 for eight of ten sites for \ac{din} (seven increasing, one decreasing), eight sites for ammonium (six increasing, two decreasing), and six sites for nitrite/nitrate (five increasing, one decreasing).  Decreasing trends were more common for the observed data from 1996-2013.  Eight sites had significant trends for \ac{din} (four increasing, four decreasing), seven sites for ammonium (five increasing, two decreasing), and eight sites for nitrite/nitrate (four increasing, four decreasing). P8 had a relatively large decrease in ammonium ($-8.3$\% change per year) for the second annual period compared to all other sites, which is described in detail in the next section. Trends by season were similar such that increases were generally observed in all seasons from 1976-1995 (Figure S1) and decreases were observed for 1996-2013 (Figure S2).  Trends for the seasonal comparisons were noisier and significant changes were less common compared to the annual comparisons.

@@

Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions.  Our results demonstrated the potential to misinterpret trends if flow effects are not considered, where the misinterpretation could vary from a simple change in the magnitude and significance of a trends, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}). A more comprehensive evaluation of flow in the Delta demonstrated that flow contributions of different end members vary considerably over time at each station\cite{Novick15}.  For example, flow at MD10 represents a changing percentage by season of inputs from the Sacramento, San Joaquin, Cosumnes, Mokelumne rivers, and agricultural returns.   For simplicity, water quality observations in our analyses were matched with large-scale drivers of flow into the Delta where most sites were matched to Sacramento or San Joaquin daily flow estimates.  Given that substantial differences with flow-normalized results were apparent from relatively coarse estimates of flow contributions, more precise differences could be obtained by considering the influence of multiple flow components at each location. Output from the Dayflow software program\cite{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description of flow-normalized trends that considers changing contributions over time.

@@

Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources (e.g., watershed agricultural load, sediment flux, etc.)\cite{Cornwell14,Novick15}.  Several \acp{wwtp} in the Delta have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate. The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 (\cref{fig:delt_map})\cite{Jabusch16}, which provides a valuable opportunity to assess how nutrient or nutrient-related source controls and water management actions have changed ambient concentrations downstream. A modal response of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
\item a shift in the ratio of the components of \ac{din} from the \ac{wwtp} before/after upgrade, and
\item a flow-normalized annual trend at P8 to show a change concurrent with \ac{wwtp} upgrades.
\end{inparaenum}

Effluent concentrations measured from 2003 to 2009 from the Stockton \ac{wwtp} showed a gradual reduction in ammonium concentration relative to total \ac{din} (Figure S3).  Ammonium and nitrate concentrations were comparable prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade, with much smaller percentages from ammonium and nitrite. As expected, flow-normalized nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}a), with ammonium showing an increase from 1976 followed by a large reduction in the 2000s.  Interestingly, nitrite/nitrate concentrations also showed a similar but less dramatic decrease despite an increase in the \ac{wwtp} effluent concentrations following the upgrade.  Percent changes from seasonal Kendall tests on flow-normalized results showed that both nitrogen species increased prior to \ac{wwtp} upgrades ($2$\% per year for nitrite/nitrate, $2.8$\% for ammonium), followed by decreases after upgrades ($-1.9$\% for nitrite/nitrate, $-16.6$\% for ammonium, \cref{tab:p8chg}).  Seasonally, increases prior to upgrades were highest in the summer for nitrite/nitrate ($2.4$\%) and in the fall for ammonium ($4.9$\%). Similarly, seasonal reductions post-upgrade were largest in the summer for nitrite/nitrate ($-4.3$\%) and largest for ammonium in the winter ($-26.7$\%).

Relationships of nitrite/nitrate with flow described by \ac{wrtds} showed an inverse flow and concentration dynamic with flushing or dilution at higher flow (\cref{fig:p8trnds}b).  Seasonal variation was even more apparent for ammonium, although both nitrite/nitrate and ammonium typically had the highest concentrations at low flow in the winter (January).  Additionally, strength of the flow/nutrient relationship changed between years. Nitrite/nitrate typically had the strongest relationship with flow later in the time series (i.e., larger negative slope), whereas ammonium had the strongest relationship with flow around 2000 in January.  Using \ac{wrtds}, an empirical link is created between upstream changes and observed effects downstream that is characterized by differences in analytes between years and season. A general conclusion is that ammonium reductions were concurrent with \ac{wwtp} upgrades, but the reduction was most apparent at low-flow in January.  These dynamics are difficult to characterize from the observed time series, and further, results from \ac{wrtds} can be used to develop additional hypotheses of factors that influence nutrient concentrations at P8. For example, estimated ammonium concentrations in July were low for all flow levels which suggests either nitrogen inputs were low in the summer or nitrogen was available and uptake by primary consumers was high. Seasonal patterns in the relationship between flow and nitrite/nitrate were not as dramatic as compared to ammonium, and in particular, low-flow events in July were generally associated with higher concentrations.  This could suggest that ammonium concentrations at P8 are driving phytoplankton production at low flow during warmer months, and not nitrite/nitrate given the higher estimated concentrations in July at low flow. As such, these simple observations from \ac{wrtds} provide quantitative support of cause and effect mechanisms of nutrient impacts on potentially adverse environmental conditions as they relate to nutrient-related source controls upstream.

@@

Invasion of the upper \ac{sfe} by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused severe changes in phytoplankton abundance and species composition.  Reduction in phytoplankton biomass has altered trophic networks in the upper \ac{sfe} and is considered an important mechanism in the decline of the protected delta smelt (\textit{Hypomesus transpacificus}) and other important fisheries \cite{Feyrer03,MacNally10}.  Changes in the physical environment have also occurred with the most notable invasion effect being increased water clarity following a reduction of phytoplankton by biofiltration \cite{MacNally10}. The clams are halophilic such that drought years are generally correlated with an increase in biomass and further upstream invasion of the species \cite{Parchaso02,Cloern12b}. We hypothesized that \ac{wrtds} models applied to water quality observations in the upper estuary would show \begin{inparaenum}[1\upshape)]
\item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, and
\item variation in the chlorophyll/clam relationship through indirect or direct controls of flow.
\end{inparaenum}
The application of \ac{wrtds} to water quality observations at station D7 in Suisun Bay and comparison with clam abundance and biomass data (see Ref~\citenum{Crauder16}) provides an approach to assess the competing effects of climate variability, hydrology, and ecology on ambient conditions.

Results from \ac{wrtds} demonstrated complex relationships between clam abundance and chlorophyll concentrations, which were further affected by flow changes over time (\cref{fig:clmchl}).  Invasion in the 1980s showed a clear displacement of the native \textit{Corbicula fluminea} with establishment of \textit{P. amurensis} (\cref{fig:clmchl}a), where biomass of the latter was negatively associated with flow from the Sacramento river (\Cref{fig:clmchl}b).  The increase in clam abundance was associated with a notable decrease in annually-averaged \ac{chla} from \ac{wrtds} results (\cref{fig:clmchl}c).  A seasonal shift in the flow-normalized results was also observed such that \ac{chla} concentrations were generally highest in July/August prior to invasion, whereas a spring maximum in April was more common in recent years (\cref{fig:clmchl}f).  An increase in annually-averaged silicon dioxide (\cref{fig:clmchl}e) was coincident with the \ac{chla} decrease, with the largest increases occuring in August (\cref{fig:clmchl}g).  These relationships suggest that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.  This supports past research that showed a decrease in silica uptake by diatoms following invasion.\cite{Cloern96,Kimmerer05} Further, \ac{din} trends were similar to silicon-dioxide in both annual and seasonal changes (compare Figures \ref{fig:clmchl}e and \ref{fig:clmchl}e with \ref{fig:clmchl}d and \ref{fig:clmchl}g) such that an increase in both nutrients earlier in the time series corresponded with the decrease in \ac{chla}.  Overall, these results suggest the increase in nutrient concentrations are from decreased uptake by phytoplankton in Suisun Bay.

The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}g) with lower \ac{chla} associated with higher biomass, confirming results from earlier studies.\cite{Alpine92,Thompson08}.  However, the effect of flow on both clams and phytoplankton as a top-down or bottom-up control changed throughout the time series.  The \ac{chla}/flow relationship showed that increasing flow (decreasing salinity) was associated with a slight increase in \ac{chla} followed by a decrease early in the time series (\cref{fig:clmchl}j), whereas overall \ac{chla} was lower but a positive association with flow (negative with salinity) was observed later in the time series. In the absence of benthic grazing prior to invasion, this dynamic suggests that \ac{chla} production may be limited at low flow as less nutrients are exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the larger bay. Following clam invasion, \ac{chla} concentrations were reduced by grazing but showed a positive and monotonic relationship with increasing flow. The increase in clam abundance was concurrent with decline in \ac{chla} concentration, although variation in abundance between years was also observed.  Clam abundance was reduced during high flow years in the late 1990s, 2006, and 2011 (\ref{fig:clmchl}a). In the same years, \ac{wrtds} predictions for \ac{chla} were higher than the flow-normalized component (\cref{fig:clmchl}c), which further suggests a link between increased flow and phytoplankton production.  As such, \ac{chla} production in early years is directly related to flow, whereas the relationship with flow in later years is indirect as increased flow reduces clam abundance and releases phytoplankton from benthic grazing pressure. These relationships have been suggested by others\cite{Alpine92,Parchaso02,Jassby08}, although the precise mechanism demonstrated by \ac{wrtds} provides a quantitative description of factors that drive water quality in the Delta.

@@

We thank the staff of the San Francisco Estuary Institute and the Delta Regional Monitoring Program. We thank Larry Harding for providing comments on an earlier draft. This study was funded, reviewed, and approved for publication by the US EPA, National Health and Environmental Efects Research Laboratory. The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the US EPA.

>>>>

Nutrient time series from 1976 to 2013 were obtained for ten discrete sampling stations (\cref{fig:delt_map}, \url{http://water.ca.gov/bdma/meta/Discrete/data.cfm})\cite{IEP13}.  Stations were grouped by location in the study area for comparison: peripheral Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; interior Delta stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations were chosen based on continuity of the water quality time series and significance of their geographic location for understanding regional trends.  Time series were complete for all stations except for an approximate ten year gap from 1996-2014 for D19.  Data were minimally processed, with the exception of averaging replicates that occurred on the same day.  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two). Less than 3\% of all observations were left-censored, although variation was observed between analytes and location.  The ammonium time series had the most censored observations at sites C10 (25.4\% of all observations), MD10 (18.1\%), D28 (17.8\%), D19 (12\%), and D7 (7.9\%).

Daily flow estimates for the Delta region were obtained from the Dayflow software program \cite{IEP16}. The \ac{wrtds} models described below require a matched flow record with the appropriate station to evaluate nutrient trends. Given the complexity of inflows and connectivity of the system, only the inflow estimates from the Sacramento and San Joaquin rivers were used as measures of freshwater influence at each station.  Initial analyses indicated that model fit was not significantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series.  As such, the Sacramento daily flow time series was used to account for flow effects at C3, D19, D26, D28, and MD10, and the San Joaquin time series was used for C10 and P8 based on station proximity to each inflow.  Salinity observations at D4, D6, and D7 in Suisun Bay were used as more appropriate measures of freshwater variation, given the stronger tidal influence at these stations.  Salinity has been used as a tracer of freshwater influence for the application of \ac{wrtds} models in tidal waters \cite{Beck15}.

@@

The observed time series for the ten Delta - Suisun Bay stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).  \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.  However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick15}.  By location, observed concentrations of \ac{din} for the entire time series were higher on average for the peripheral stations (C3, C10, MD10, P8; mean $\pm$ s.e.: $1.04$$\pm$$0.03$ \mgl) and similar for the interior (D19, D26, D28, $0.43$$\pm$$0.01$) and Suisun Bay stations (D4, D6, D7, $0.44$$\pm$$0.01$).  Average concentrations were highest at P8 ($1.63$$\pm$$0.05$ \mgl) and lowest at C3 ($0.4$$\pm$$0.01$) for \ac{din}, highest at P8 ($0.28$$\pm$$0.02$) and lowest at D28 ($0.05$$\pm$$0.003$) for ammonium, and highest at C10 ($1.4$$\pm$$0.04$) and lowest at C3 ($0.15$$\pm$$0.004$) for nitrite/nitrate. Mean observed concentrations were also higher later in the time series for all analytes.  For example, average \ac{din} across all stations was $0.61$$\pm$$0.01$ \mgl for 1976-1995, compared to $0.7$$\pm$$0.01$ for 1996-2013. Seasonal changes across all years showed that nitrogen concentrations were generally lower in the summer and higher in the winter, although observed patterns were inconsistent between sites.  For example, site MD10 had distinct seasonal spikes for elevated \ac{din} in the winter, whereas other stations had less prominent seasonal maxima (e.g., C3, D7, \cref{fig:obsdat}).

@@

Application of seasonal Kendall tests to evaluate trends in observed data provided information on the direction, magnitude, and statistical significance of changes between years. Trends estimated for 1976-1995 and 1996-2013 varied considerably between sites and analytes (\cref{fig:trndcomp1}). Significant trends were observed from 1976-1995 for eight of ten sites for \ac{din} (seven increasing, one decreasing), eight sites for ammonium (six increasing, two decreasing), and six sites for nitrite/nitrate (five increasing, one decreasing).  Decreasing trends were more common for the observed data from 1996-2013.  Eight sites had significant trends for \ac{din} (four increasing, four decreasing), seven sites for ammonium (five increasing, two decreasing), and eight sites for nitrite/nitrate (four increasing, four decreasing). P8 had a relatively large decrease in ammonium ($-8.3$\% change per year) for the second annual period compared to all other sites (see next section). Trends by season were similar such that increases were generally observed in all seasons from 1976-1995 (Figure S1) and decreases were observed for 1996-2013 (Figure S2).  Trends for the seasonal comparisons were noisier and significant changes were less common compared to the annual comparisons.

@@

Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions.  Our results demonstrated the potential to misinterpret trends if flow effects are not considered, where the misinterpretation could vary from a simple change in the magnitude and significance of a trend, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}). A more comprehensive evaluation of flow in the Delta demonstrated that flow contributions of different end members vary considerably over time at each station\cite{Novick15}.  For example, flow at MD10 represents a changing percentage by season of inputs from the Sacramento, San Joaquin, Cosumnes, Mokelumne rivers, and agricultural returns.   For simplicity, water quality observations in our analyses were matched with large-scale drivers of flow into the Delta where most sites were matched to Sacramento or San Joaquin daily flow estimates.  Given that substantial differences with flow-normalized results were apparent from relatively coarse estimates of flow contributions, more precise differences could be obtained by considering the influence of multiple flow components at each location. Output from the Dayflow software program\cite{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description of flow-normalized trends that considers changing contributions over time.

@@

Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources \cite{Cornwell14,Novick15}.  Several \acp{wwtp} in the Delta have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate. The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \cite{Jabusch16}, which provides a valuable opportunity to assess how nutrient or nutrient-related source controls and water management actions have changed ambient concentrations downstream. A modal response of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
\item a shift in the ratio of the components of \ac{din} from the \ac{wwtp} before/after upgrade, and
\item a flow-normalized annual trend at P8 to show a change concurrent with \ac{wwtp} upgrades.
\end{inparaenum}

Effluent measured from 2003 to 2009 from the Stockton \ac{wwtp} had a gradual reduction in ammonium concentration relative to total \ac{din} (Figure S3).  Ammonium and nitrate concentrations were comparable prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade, with much smaller percentages from ammonium and nitrite. As expected, flow-normalized nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}a), with ammonium showing an increase from 1976 followed by a large reduction in the 2000s.  Interestingly, nitrite/nitrate concentrations also showed a similar but less dramatic decrease despite an increase in the \ac{wwtp} effluent concentrations following the upgrade.  Percent changes from seasonal Kendall tests on flow-normalized results showed that both nitrogen species increased prior to \ac{wwtp} upgrades ($2$\% per year for nitrite/nitrate, $2.8$\% for ammonium), followed by decreases after upgrades ($-1.9$\% for nitrite/nitrate, $-16.6$\% for ammonium, \cref{tab:p8chg}).  Seasonally, increases prior to upgrades were highest in the summer for nitrite/nitrate ($2.4$\%) and in the fall for ammonium ($4.9$\%). Similarly, seasonal reductions post-upgrade were largest in the summer for nitrite/nitrate ($-4.3$\%) and largest for ammonium in the winter ($-26.7$\%).

Relationships of nitrite/nitrate with flow described by \ac{wrtds} showed an inverse flow and concentration dynamic with flushing or dilution at higher flow (\cref{fig:p8trnds}b).  Seasonal variation was even more apparent for ammonium, although both nitrite/nitrate and ammonium typically had the highest concentrations at low flow in the winter (January).  Additionally, strength of the flow/nutrient relationship changed between years. Nitrite/nitrate typically had the strongest relationship with flow later in the time series (i.e., larger negative slope), whereas ammonium had the strongest relationship with flow around 2000 in January. A general conclusion is that ammonium reductions were concurrent with \ac{wwtp} upgrades, but the reduction was most apparent at low-flow in January.  These dynamics are difficult to characterize from the observed time series, and further, results from \ac{wrtds} can be used to develop additional hypotheses of factors that influence nutrient concentrations at P8. For example, estimated ammonium concentrations in July were low for all flow levels which suggests either nitrogen inputs were low in the summer or nitrogen was available and uptake by primary consumers was high. Seasonal patterns in the relationship between flow and nitrite/nitrate were not as dramatic as compared to ammonium, and in particular, low-flow events in July were associated with higher concentrations.  This could suggest that ammonium concentrations at P8 are driving phytoplankton production at low flow during warmer months, and not nitrite/nitrate given the higher estimated concentrations in July at low flow. As such, these simple observations provide quantitative support of cause and effect mechanisms of nutrient impacts on potentially adverse environmental conditions as they relate to nutrient-related source controls upstream.

@@

Invasion of the upper \ac{sfe} by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused severe changes in phytoplankton abundance and species composition.  Reduction in phytoplankton biomass has altered trophic networks in the upper \ac{sfe} and is considered an important mechanism in the decline of the protected delta smelt (\textit{Hypomesus transpacificus}) and other important fisheries \cite{Feyrer03,MacNally10}.  Changes in the physical environment have also occurred, such as increased water clarity following a reduction of phytoplankton by biofiltration \cite{MacNally10}. The clams are halophilic such that drought years are correlated with an increase in biomass and further upstream invasion of the species \cite{Parchaso02,Cloern12b}. We hypothesized that results from \ac{wrtds} models would show \begin{inparaenum}[1\upshape)]
\item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, and
\item variation in the chlorophyll/clam relationship through indirect or direct controls of flow.
\end{inparaenum}

Invasion in the 1980s showed a clear displacement of the native \textit{Corbicula fluminea} with establishment of \textit{P. amurensis} (\cref{fig:clmchl}a), where biomass of the latter was negatively associated with flow from the Sacramento river (\Cref{fig:clmchl}b).  The increase in clam abundance was associated with a notable decrease in annually-averaged \ac{chla} from \ac{wrtds} results (\cref{fig:clmchl}c).  A seasonal shift in the flow-normalized results was also observed such that \ac{chla} concentrations were generally highest in July/August prior to invasion, whereas a spring maximum in April was more common in recent years (\cref{fig:clmchl}f).  An increase in annually-averaged silicon dioxide (\cref{fig:clmchl}e) was coincident with the \ac{chla} decrease, with the largest increases occuring in August (\cref{fig:clmchl}g).  These relationships suggest that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.  This supports past research that showed a decrease in silica uptake by diatoms following invasion.\cite{Cloern96,Kimmerer05} Further, \ac{din} trends were similar to silicon-dioxide in both annual and seasonal changes (compare Figures \ref{fig:clmchl}e and \ref{fig:clmchl}e with \ref{fig:clmchl}d and \ref{fig:clmchl}g) such that an increase in both nutrients earlier in the time series corresponded with the decrease in \ac{chla}.  Overall, these results suggest the increase in nutrient concentrations are from decreased uptake by phytoplankton in Suisun Bay.

The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}g), with lower \ac{chla} associated with higher biomass, confirming results from earlier studies.\cite{Alpine92,Thompson08}.  However, the effect of flow on both clams and phytoplankton as a top-down or bottom-up control changed throughout the time series.  The \ac{chla}/flow relationship showed that increasing flow (decreasing salinity) was associated with a slight increase in \ac{chla} followed by a decrease early in the time series (\cref{fig:clmchl}j), whereas overall \ac{chla} was lower but a positive association with flow (negative with salinity) was observed later in the time series. In the absence of benthic grazing prior to invasion, this dynamic suggests that \ac{chla} production may be limited at low flow as less nutrients are exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the larger bay. Following clam invasion, \ac{chla} concentrations were reduced by grazing but showed a positive and monotonic relationship with increasing flow. The increase in clam abundance was concurrent with decline in \ac{chla} concentration, although variation in abundance between years was also observed.  Clam abundance was reduced during high flow years in the late 1990s, 2006, and 2011 (\ref{fig:clmchl}a). In the same years, \ac{wrtds} predictions for \ac{chla} were higher than the flow-normalized component (\cref{fig:clmchl}c), which further suggests a link between increased flow and phytoplankton production.  As such, \ac{chla} production in early years is directly related to flow, whereas the relationship with flow in later years is indirect as increased flow reduces clam abundance and releases phytoplankton from benthic grazing pressure. These relationships have been suggested by others\cite{Alpine92,Parchaso02,Jassby08}, although the precise mechanism demonstrated by \ac{wrtds} provides a quantitative description of factors that drive water quality in the Delta.

@@

We thank the staff of the San Francisco Estuary Institute and the Delta Regional Monitoring Program. We thank Larry Harding for providing comments on an earlier draft. This study was reviewed and approved for publication by the US EPA, National Health and Environmental Efects Research Laboratory. The authors declare no competing financial interest. The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the US EPA.

### 2017-05-02 supporting.Rnw
slight shortening

% supplementary material
\section*{Supplementary Information}

@@

\noindent The following files are available free of charge: \cref{fig:trndcomp2,fig:trndcomp3,fig:stock}, \cref{tab:trndsann,tab:trndsbef,tab:trndsaft}

@@

# table stuff
cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and annual aggregations.  Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic). Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.'
foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
stats <- tab$Site_Code

@@

# table stuff
cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1976-1995. Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic). Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period. Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
stats <- tab$Site_Code

@@

# table stuff
cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1996-2013. Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic). Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period. Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.'
foot.val<-'\\footnotesize *$p<0.05$; **$p<0.005$'
vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
stats <- tab$Site_Code

>>>>

% supplementary material
\section*{Supporting Information}

@@

\noindent The following files are available free of charge: \cref{fig:trndcomp2,fig:trndcomp3,fig:stock}, \cref{tab:trndsann,tab:trndsbef,tab:trndsaft} (7 pages)

@@

# table stuff
cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and annual aggregations'
foot.val<-'\\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic). Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period. *$p<0.05$; **$p<0.005$'
vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
stats <- tab$Site_Code

@@

# table stuff
cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1976-1995'
foot.val<-'\\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic). Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period. Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF. *$p<0.05$; **$p<0.005$'
vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
stats <- tab$Site_Code

@@

# table stuff
cap.val <- 'Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1996-2013'
foot.val<-'\\footnotesize Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic). Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period. Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF. *$p<0.05$; **$p<0.005$'
vars <- c('DIN', 'NH$_{4}^{+}$', 'NO$_{2}^{-}$/NO$_{3}^{2-}$')
stats <- tab$Site_Code
